Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-29 parallel-tractability-J-v9/parallel-tractability-J.tex
splitting the main file for better organization

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep \texttt{NC} complexity
\end{keyword}

@@

\section{Introduction}
\label{sec:introduction}

The Web Ontology Language (OWL)\footnote{The latest version is OWL 2: http://www.w3.org/TR/owl2-overview/}
is an important standard for ontology language in the Semantic Web and knowledge-based applications.
In these applications, \emph{materialization}, which is the reasoning task of computing all implicit
facts that follow from a given ontology \cite{handbook}, plays an important role. By means of it, developers and users
can optimize query answering, ontology diagnosis and debugging. Since huge amounts of semantic data
are being generated at an increasing pace by sensor networks, government authorities and social
media \cite{LehmbergRMB16,MeuselBP15}, it is challenging to conduct materialization on such large-scale ontologies efficiently.

There has been work that proposes parallel reasoning algorithms and employs parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balance workload,
data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, in order to make reasoning tasks more efficient
in parallel, several work utilizes different kinds of high-performance computing platforms, like supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that, some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, we show in Section~\ref{sec:evaluation} that, there do exist some ontologies, for which materialization cannot
always be improved by parallelism. The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class $\texttt{NC}$ \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the \texttt{NC} class, one can give an \emph{\texttt{NC} algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above \texttt{NC} algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

\section{Background Knowledge}
\label{sec:background}

In this section, we introduce some basic notions that are needed to introduce our approach.

\subsection{Datalog}

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,...,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,...,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,...,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,...,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

\subsection{DHL and DL-Lite}

In what follows, we use $\textbf{CN}$, $\textbf{RN}$ and $\textbf{IN}$
to denote three disjoint countably
infinite sets of \emph{concept names, role names}, and \emph{individual names} respectively.
The set of roles is defined as $\textbf{R}:=\textbf{RN}\cup\{R^-|R\in\textbf{RN}\}$
where $R^-$ is the \emph{inverse role} of $R$.
For ease of discussion, we focus on the \emph{simple forms} of axioms shown
in the left column of Table~1. These simple forms can be obtained by using
well-known \emph{structure transformation} techniques \cite{KrotzschRH07,Kazakov09}.

DHL (short for \emph{description horn logic}) \cite{GrosofHVD03} is introduced as an
intersection of description logic (DL) and datalog in terms of expressivity.
We define a DHL ontology $\mathcal{O}$ as a triple:
$\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$, where
$\mathcal{T}$ denotes the TBox containing axioms of the forms (T1-T3);
$\mathcal{R}$ is the RBox that is a set of axioms of the forms (R1-R3);
$\mathcal{A}$ is the ABox containing \emph{assertions} of the forms (A1) and (A2).
In an axiom of either of  the forms (T1-T3 and R1-R3), concepts $A_{(i)}$ and $B$ are either
concept names, \emph{top concept} ($\top$) or \emph{bottom concept} ($\bot$); $R$ and $S_{(i)}$
are roles in $\textbf{R}$.
For an axiom of the form $A\sqsubseteq\forall R.B$ that is also allowed in DHL, we only consider its
equivalent form $\exists R^-.A\sqsubseteq B$.

DHL is related with other ontology languages.
First, DHL is essentially a fragment of the description logic Horn-$\mathcal{SHOIQ}$ with
disallowing \emph{nominal}, \emph{number restriction} and
right-hand \emph{existential restriction} ($A\sqsubseteq\exists R.B$).
Second, the expressivity of DHL covers that of RDFS to some extent \cite{GrosofHVD03}.
Reasoning with RDFS ontologies is \texttt{NP}-complete \cite{Horst05}
and, thus, is not parallelly tractable.
However, by applying some simplifications and restrictions, RDFS ontologies can be
expressed in DHL \cite{GrosofHVD03} that has \texttt{PTime}-complete complexity for
materialization.

\begin{table}
\begin{center}
\begin{tabular}{lcc}
\multicolumn{3}{c}{\textbf{Table 1: Axioms and corresponding datalog rules}}\\
\hline
&~~~~~~~~~~~~~~~~~~~~~~~~Axioms~~~~~~~~~~~~~~~~~~~~~~~~&~~~~~~~~~~~~~~~~~~~~~~~~Datalog Rules~~~~~~~~~~~~~~~~~~~~~~~~\\
\hline
\hline
(T1)& $A\sqsubseteq B$& $A(x)\rightarrow B(x)$\\
(T2)& $A_1\sqcap A_2\sqsubseteq B$& $A_1(x), A_2(x)\rightarrow B(x)$\\
(T3)& $\exists R.A\sqsubseteq B$& $R(x,y), A(y)\rightarrow B(x)$\\
(T4)& $A\sqsubseteq\exists R$& $A(x)\rightarrow R(x, o_{R}^A)$\\

\hline
(R1)& $S\sqsubseteq R$& $S(x,y)\rightarrow R(x,y)$\\
(R2)& $S\sqsubseteq R^-$& $S(x,y)\rightarrow R(y,x)$\\
(R3)& $R\circ R\sqsubseteq R$& $R(x,y), R(y,z)\rightarrow R(x,z)$\\
(R4)& $R_1\circ R_2\sqsubseteq R$&$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$\\

\hline
(A1)& $A(a)$& $A(a)$\\
(A2)& $R(a,b)$& $R(a,b)$\\
\hline
\end{tabular}
\label{tab:dhl}
\end{center}
\end{table}

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed to datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of the form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

DL-Lite is a group of ontology languages designed for highly-efficient
query answering through knowledge
and underpins the OWL profile OWL QL \cite{CalvaneseGLLR07}.
DL-Lite$_{core}$ and DL-Lite$_{\mathcal{R}}$ are the two basic fragments
in DL-Lite.
DL-Lite$_{core}$ requires that an TBox contains axioms of the forms (T1) $A\sqsubseteq B$,
(T2) $A_1\sqcap A_2\sqsubseteq B$ where $B\equiv\bot$, (T3) $\exists R.A\sqsubseteq B$
where $A\equiv\top$ or $B\equiv\bot$, and (T4) $A\sqsubseteq\exists R$,
an ABox contains \emph{assertions} of the forms (A1) and (A2), and
RBox is not involved. DL-Lite$_{\mathcal{R}}$ is an extension of DL-Lite$_{core}$,
which allows involving RBoxes that contain axioms of the forms (R1-R2).
If not specially specified, \emph{a DL-Lite ontology denotes a DL-Lite$_{\mathcal{R}}$ ontology
in the following paragraphs}.

\subsection{Ontology Materialization via Datalog Programs}

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is a free variable.
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we also use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules transformed from the axioms in $\mathcal{T}$ and $\mathcal{R}$,
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which is initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}.

\subsection{The $\texttt{NC}$ Complexity}

The parallel complexity class $\texttt{NC}$, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the $\texttt{NC}$ class
can be solved in poly-logarithmic time on a parallel machine with a polynomial number of processors.
We also say that an $\texttt{NC}$ problem can be solved in \emph{parallel poly-logarithmic time}.
Although the $\texttt{NC}$ complexity is a theoretical analysis tool,
it has been shown that many $\texttt{NC}$ problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the $\texttt{NC}$ problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The \texttt{NC} complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in $\texttt{NC}$.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

\section{Parallel Tractability of Datalog Programs}
\label{sec:ptclass}

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any datalog program $\langle R, \textbf{I}\rangle$ the rule
set $R$ is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an \texttt{NC} algorithm)}.
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

\subsection{Parallel Tractable Classes}

We first give the following definition for a parallel tractable class of datalog programs
where an \texttt{NC} algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs,
we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an \texttt{NC} algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our $\texttt{NC}$ algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an \texttt{NC} version
and identify the target \texttt{PTD} class.

\subsection{Materialization Graph}

In order to give a parallel materialization algorithm,
we introduce the notion of \emph{materialization graph}.
It makes the analysis of the given algorithm convenient.

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$;
\end{enumerate}
~~$\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph, the nodes whose in-degree is 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes whose in-degree is 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-mg.eps}
\caption{An example of a materialization graph.}
\label{fig:mg}
\end{center}
\end{figure}

We say that a materialization graph $\mathcal{G}$ is a \emph{complete materialization graph}
when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
The set of nodes in a complete materialization graph is actually the result
of materialization.
Thus, the procedure of materialization can be transformed to the construction of
a complete materialization graph.
We pay our attention to complete materialization graphs and do not distinguish it to the
notion `materialization graph'.
It should also be noted that there may exist several materialization graphs for a datalog program.

\subsection{A Basic Parallel Algorithm}

In this part, we propose a parallel algorithm
(denoted by $\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.\\

\noindent\texttt{Algorithm $\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Recall that $P^*$ denotes the ground instantiation of $P$,
which consists of all possible ground instantiations of rules in $R$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.\footnote{This might not be practically feasible,
but we focus on a theoretical analysis here.}
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,\footnote{Suppose that each processor
    can use $O(1)$ time units to access the state of ground atoms, i.e., whether this ground
    atom has been added to the materialization graph. This can be implemented by maintaining an
    index of polynomial size.}
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

\begin{example}
We consider the datalog program $P_{ex_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{bsc}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{ex_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{ex_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{ex_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{ex_1}$.
Algorithm~$\mathsf{A}_{bsc}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{ex_1}$ (\ref{alg1:halt}).
\end{example}

Lemma~\ref{lemma:a1} shows the correctness of Algorithm~$\mathsf{A}_{bsc}$ and that, for any datalog program $P$,
Algorithm~$\mathsf{A}_{bsc}$ always constructs a materialization graph with the minimum depth among all the
materialization graphs of $P$. The detailed proofs of Lemma~\ref{lemma:a1} and other lemmas and theorems can be found
in the appendix.

\begin{lemma}
\label{lemma:a1}
Given a datalog program $P=\langle R, \textbf{I}\rangle$, we have
\begin{enumerate}[leftmargin=4ex]
\item Algorithm~$\mathsf{A}_{bsc}$ halts and returns a materialization graph $\mathcal{G}$ of $P$;
\item $\mathcal{G}$ has the the minimum depth among all the materialization graphs of $P$.
\end{enumerate}
\end{lemma}

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{bsc}$,
its depth equals the depth of the stage.
\end{proof}

We now discuss how Algorithm~$\mathsf{A}_{bsc}$ can be restricted to an \texttt{NC} version.
(\uppercase\expandafter{\romannumeral1}) Since Algorithm~$\mathsf{A}_{bsc}$ does not introduce
new constants and each predicate has a constant arity,
one can check that $|P^*|$ is polynomial in the size of $P$. This also means that
the number of processors is polynomially bounded.
(\uppercase\expandafter{\romannumeral2}) The computing time of \ref{alg1:addFacts}
and \ref{alg1:halt} occupies constant time units because of parallelism.
(\uppercase\expandafter{\romannumeral3}) The main computation part in Algorithm~$\mathsf{A}_{bsc}$ is the
iteration of \ref{alg1:updateG}.
In each iteration of \ref{alg1:updateG}, all processors work independently from each other.
Thus, in theory, \ref{alg1:updateG} costs one time unit.
The whole computing time turns out to be bounded by the number of iterations of \ref{alg1:updateG}. (\uppercase\expandafter{\romannumeral4})
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
The input of $\psi$ is the size of $P$ and the output is a non-negative integer.
Based on (\uppercase\expandafter{\romannumeral1}, \uppercase\expandafter{\romannumeral2},
\uppercase\expandafter{\romannumeral3}, \uppercase\expandafter{\romannumeral4}),
for any datalog program $P$, if we use $\psi(|P|)$ to
bound the number of iterations of \ref{alg1:updateG}, then Algorithm~$\mathsf{A}_{bsc}$
is an \texttt{NC} algorithm, denoted by $\mathsf{A}_{bsc}^{\psi}$ (this symbol also
means that $\psi$ occurs as a configuration for Algorithm~$\mathsf{A}_{bsc}$).

Based on $\mathsf{A}_{bsc}^{\psi}$, we can identify a class of datalog programs
$\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ such that all the datalog programs in it can be handled
by $\mathsf{A}_{bsc}^{\psi}$.
It is obvious that $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ is a \texttt{PTD} class.
We use the following theorem to further show that this class can be captured based on
the materialization graphs of the datalog programs in $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$.

\begin{theorem}\label{theorem:a1}
For any datalog program $P$, $P\in\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ iff $P$ has a
materialization graph whose depth is upper-bounded by $\psi(|P|)$.
\end{theorem}

\begin{proof}[Proof sketch]
We can first prove that the number of iterations
of \ref{alg1:updateG} is actually the depth of the constructed materialization
graph. This theorem then follows by considering
Lemma~\ref{lemma:a1}.
\end{proof}

The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the
rather simple datalog program $P_{ex_1}$ in Example~\ref{exp:mg}.
The graph $\mathcal{G}_{ex_1}$ in Figure~\ref{fig:mg} is the unique materialization graph of $P_{ex_1}$.
One can also check that \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$.
This means that the depth of $\mathcal{G}_{ex_1}$ is linearly bounded by $k$.
On the other hand, the size of $P_{ex_1}$ is a polynomial of the integer $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph of $P_{ex_1}$.
However there indeed exists an \texttt{NC} algorithm that can handle $P_{ex_1}$.
We discuss this in the next part.

\subsection{Optimizing Algorithm~$\mathsf{A}_{bsc}$ via the Single-Way Derivability}

In this part, we optimize Algorithm~$\mathsf{A}_{bsc}$ such that $P_{ex_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{bsc}$,
we can identify another \texttt{PTD} class.

We discuss our optimization based on a specific case in Example~\ref{exp:general}.
We find that, in this kind of case, the construction of a materialization graph can be accelerated.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-general.eps}
\caption{A partial materialization graph.}
\label{fig:general}
\end{center}
\end{figure}

\begin{example}\label{exp:general}
Consider a snapshot of Algorithm~$\mathsf{A}_{bsc}$ in
Figure~\ref{fig:general}. A materialization graph $\mathcal{G}$
is being constructed for some datalog program $\langle R, \textbf{I}\rangle$.
The nodes in the dashed box denote the ones that have been added to $\mathcal{G}$.
In this snapshot, $v_0$ has been newly added
to $\mathcal{G}$ in the $l^{th}$ ($l\geq 1$) iteration.
Each of the nodes $v_i$ ($1\leq i\leq k$) has almost one parent node being not in $\mathcal{G}$,
while $v'$ has two parent nodes being not in $\mathcal{G}$.
All of the nodes $v_i$ ($1\leq i\leq k$) and $v'$ would be added to $\mathcal{G}$
afterwards.
\end{example}

In Example~\ref{exp:general}, $v_k$ would be added to $\mathcal{G}$ after \emph{at least} $k$
iterations by performing Algorithm~$\mathsf{A}_{bsc}$.
We can check that each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) in this stage follows the condition:
\emph{if the parent node $v_{i-1}$ is derivable, the child node $v_i$ has to be derivable} (this
is because that, for node $v_i$, node $v_{i-1}$ is the only parent node that has not been added to $\mathcal{G}$).
We call this condition \emph{the single-way derivability condition}, which intuitively says that the derivability of a
node only depends on one of its parent nodes.
Observe that node $v_k$ is reachable from node $v_0$ through the path $\tau=(v_0,v_1,...,v_k)$ where
each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) follows the single-way derivability condition;
we call such a path a \emph{single-way derivable} (SWD) path.
Further, the starting node $v_0$ in $\tau$ has been added to $\mathcal{G}$; in other words,
$v_0$ is derivable. Thus, all of the nodes $v_i$ ($1\leq i\leq k$) can be added to
$\mathcal{G}$ immediately.
Based on the above idea, we optimize Algorithm~$\mathsf{A}_{bsc}$ by checking whether such an SWD path
exits for some node $v$; if there is an SWD path for node $v$, this node can be added to $\mathcal{G}$
due to the single-way derivability condition.

Backing to Example~\ref{exp:general}, since
there exists an SWD path $(v_0,...,v_i)$ for each node $v_i$ ($1\leq i\leq k$), all of these nodes
can be added to $\mathcal{G}$ right after $v_0$. On the other hand,
node $v'$ has no SWD path, since it has two parent nodes being not in $\mathcal{G}$.

We next discuss how to determine the existence of SWD paths. Note that, SWD paths require us to
describe the reachability between two nodes. To this extent, we use a
binary transitive relation $\texttt{rch} \subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$,
e.g., \texttt{rch}$(v_1,v_2)$ means that $v_2$ is reachable from $v_1$.
In each iteration of \ref{alg1:updateG} in Algorithm~$\mathsf{A}_{bsc}$, we further compute a \texttt{rch}
relation (denoted by $S_{\textit{\tiny rch}}$) by performing the following process:

>>>>

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep NC complexity
\end{keyword}

@@

\input{introduction}

\input{background}

\input{parallelTractabilityDatalog}

\input{parallelTractabilityOntMat}

\input{optimizedAlgo}

\input{evaluation}

\input{relatedWork}

\input{conclusions}
